---
title: "Comparing Methods for Handling Imbalanced Datasets"
author: "Hamed Helisaz"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(tidyverse)
library(imbalance)
library(GGally)
library(patchwork)
library(recipes)
library(themis)
library(smotefamily)
library(UBL)
library(unbalanced)
library(devtools)
library(caret)
library(formattable)
library(kableExtra)
set.seed(456)
```


### 1. Data Preparation 💾

Loading `newthyroid1` data set from `imbalanced` library

```{r 1.1 Load Data, echo = FALSE}

data_raw <- read.csv("data/data.csv")


factor_cols <- c("gender", "hypertension","heart_disease","ever_married",
                 "work_type","Residence_type","smoking_status")


data <- data_raw %>%
  mutate(Class = ifelse(stroke == 0, "negative","positive"))%>%
  select(-stroke)%>%
  mutate(Class = factor(Class, levels = c("negative","positive")))%>%
  select(-id)%>%
  mutate(across(all_of(factor_cols),as.factor))%>%
  ## Addressing NA
  mutate(smoking_status = ifelse(is.na(smoking_status), "Others",smoking_status))%>%
  filter(if_all(everything(), ~!is.na(.)))


fontsize <- 8
basesize <- 8
```

___

### 2. Exploratory Data Analysis 🔍
#### 2.1. Initial Data Overview 📊

**Format**

A data frame with `r comma(nrow(data),0)` records and `r length(names(data))` columns; `r comma(nrow(data[data['Class'] == "positive",]),0)` comes with positive (minority) outcomes and `r comma(nrow(data[data['Class'] == "negative",]),0)` comes with negative (majority) outcomes (`r percent(nrow(data[data['Class'] == "positive",])/nrow(data),0)` imbalance). 

**Variables**

`r names(data)`

```{r 2.1 Initial Data Overview, echo = FALSE}

glimpse(data)

```


___

#### 2.2. Visualization
###### 2.2.1. Distribution of Classes

```{r 2.2 Class Distribution, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = 'center'}

ggplot(data, aes(x = Class, fill = Class)) +
  geom_bar() +
  labs(title = "Unbalanced Distribution", x = "Class", y = "Count") +
  theme_minimal(base_size = basesize) +
  scale_fill_manual(values = c("positive" = "skyblue", "negative" = "tomato"))+
  theme(axis.title.x = element_text(size = fontsize),
        axis.text.x = element_text(size = fontsize),
        axis.title.y = element_text(size = fontsize),
        axis.text.y = element_text(size = fontsize),
        plot.title = element_text(size = fontsize),
        legend.title = element_text(size = fontsize),
        legend.text = element_text(size = fontsize),
        legend.key.size =  unit(2, 'mm'),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7))

```

___

###### 2.2.2. Distribution of Predictors

```{r 2.2 Predictors Histograms, echo = FALSE, fig.width = 9, fig.height = 6, fig.align = 'center'}

numeric_predictors <- setdiff(names(data), c("Class",factor_cols))

plot_list <- list()

for (col_name in numeric_predictors) {
p <- ggplot(data, aes(x = .data[[col_name]])) +
    geom_histogram(aes(y = after_stat(density)), binwidth = NULL, fill = "lightblue", color = "black", alpha = 0.7) +
    geom_density(color = "darkblue", linewidth = 1) +
    labs(
      title = paste("Distribution of", col_name),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal()
   plot_list[[col_name]] <- p
}
   

combined_plot <- wrap_plots(plot_list, ncol = 2) 

print(combined_plot)

```

___

###### 2.2.3. Distribution of Predictors by Class - Histogram

```{r 2.2 Predictors Histograms by Class, echo = FALSE, fig.width = 9, fig.height = 6, fig.align = 'center'}


plot_list <- list()
class_colors <- c("positive" = "skyblue", "negative" = "tomato")

for (col_name in numeric_predictors) {
  p <- ggplot(data, aes(x = .data[[col_name]], fill = Class, color = Class)) +
    geom_density(alpha = 0.5, linewidth = 0.8) +
    scale_fill_manual(values = class_colors) +
    scale_color_manual(values = class_colors) +
    labs(
      title = paste("Distribution of", col_name, "by Class"),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal(base_size = fontsize -1) +
    theme(legend.position = "none")

  plot_list[[col_name]] <- p
}

combined_plot <- wrap_plots(plot_list, ncol = 2) +
                 plot_layout(guides = 'collect') &
                 theme(legend.position = 'right')

print(combined_plot)
```


###### 2.2.3. Distribution of Predictors by Class - Contingency Tables

```{r 2.2 Predictors Boxplot, echo = FALSE, results='asis'}

factor_predictors <- factor_cols  

tables_list <- list()

for (col_name in factor_predictors) {
  tbl_df <- data %>%
    group_by(!!sym(col_name), Class) %>%
    summarise(Count = n(), .groups = "drop") %>%
    group_by(!!sym(col_name)) %>%
    mutate(Percent = round(100 * Count / sum(Count), 1),
           Label = paste0(comma(Count), " (", Percent, "%)")) %>%
    select(!!sym(col_name), Class, Label) %>%
    pivot_wider(names_from = Class, values_from = Label) %>%
    mutate(across(everything(), ~replace_na(.x, "0 (0.0%)")))
  
  tables_list[[col_name]] <- kable(tbl_df, format = "html") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE, position = "left",
                  font_size = 12)
}


for (i in seq(1, length(tables_list), by = 4)) {
cat('<div style="display: flex; justify-content: flex-start; gap: 20px;">')

print(tables_list[[i]])

if ((i + 1) <= length(tables_list)) {
  print(tables_list[[i + 1]])
}

if ((i + 2) <= length(tables_list)) {
  print(tables_list[[i + 2]])
}

if ((i + 3) <= length(tables_list)) {
  print(tables_list[[i + 3]])
}

cat('</div><br>\n')
}
```

___

### 3. Methodology 🛠️

##### 3.1. Baseline Model (No Imbalance Treatment)

```{r 3.1 Baseline, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

train_control <- trainControl(method = "cv",
                              number = 5,
                              classProbs = TRUE,
                              summaryFunction = twoClassSummary,
                              savePredictions = "final")


train_index <- createDataPartition(data$Class, p = 0.7, list = FALSE)
data_train <- data[train_index, ]
data_test <- data[-train_index, ]

model_baseline <- train(Class ~ .,
                        data = data_train,
                        method = "glm",
                        trControl = train_control,
                        metric = "ROC")
                        ##  "F1", "Recall", "Sensitivity", "Specificity"

print(model_baseline)
predictions_baseline <- predict(model_baseline, newdata = data_test)

table(data_train$Class)
confusionMatrix(predictions_baseline, data_test$Class)

```

___

##### 3.2. Resampling Techniques
###### 3.2.1. Upsampling Methods - Random Over-Sampling (ROS)

```{r 3.1 ROS, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

data_train_upsampled_ros <- upSample(
  x = subset(data_train, select = -Class),
  y = data_train$Class,
  yname = "Class")

table(data_train_upsampled_ros$Class)

model_upsample_ros <- train(Class ~ .,
                            data = data_train_upsampled_ros,
                            method = "glm",
                            trControl = train_control,
                            metric = "ROC")

predictions_upsample_ros <- predict(model_upsample_ros, newdata = data_test)
confusionMatrix(predictions_upsample_ros, data_test$Class)

```

___

###### 3.2.2. Upsampling Methods -  SMOTE (Synthetic Minority Over-sampling Technique)

```{r 3.1 SMOTE, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


prepped_recipe_smote <- recipe(Class ~ ., data = data_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_smote(Class)%>%prep()

data_train_upsampled_smote <- juice(prepped_recipe)
data_test_upsampled_smote <- bake(prepped_recipe, new_data = data_test)

table(data_train_upsampled_smote$Class)


model_upsample_smote <- train(Class ~ .,
                              data = data_train_upsampled_smote,
                              method = "glm",
                              family = "binomial",
                              trControl = train_control,
                              metric = "ROC")

print(model_upsample_smote)

predictions_upsample_smote_prob <- predict(model_upsample_smote, newdata = data_test_upsampled_smote, type = "prob")

predictions_upsample_smote_class <- predict(model_upsample_smote, newdata = data_test_upsampled_smote, type = "raw")

confusionMatrix(predictions_upsample_smote_class,
                data_test$Class)

```

___

###### 3.2.3. Upsampling Methods -  SMOTE-ENN

```{r 3.1 SMOTE-ENN, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

prepped_recipe_smote_enn <- recipe(Class ~ ., data = data_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_smote(Class, over_ratio = 1)%>%prep()


data_train_upsampled_smote_enn <- juice(prepped_recipe_smote_enn)
data_test_upsampled_smote_end <- bake(prepped_recipe_smote_enn, new_data = data_test)


data_train_upsampled_smote_enn$Class <- ifelse(
  data_train_upsampled_smote_enn$Class == "negative", 0, 1)

data_train_upsampled_smote_enn <- ubENN(X = data_train_upsampled_smote_enn
                    [, setdiff(names(data_train_upsampled_smote_enn), "Class")],
                    Y =  data_train_upsampled_smote_enn$Class)

data_train_upsampled_smote_enn_final <-
  cbind(data_train_upsampled_smote_enn$X,
        data.frame(Class = factor(
          ifelse(
          data_train_upsampled_smote_enn$Y == 0, 'negative','positive'),
          level = c('negative','positive'))))

model_upsample_smote_enn <- train(Class ~ .,
                              data = data_train_upsampled_smote_enn_final,
                              method = "glm",
                              family = "binomial",
                              trControl = train_control,
                              metric = "ROC")


predictions_upsample_smote_enn_class <- predict(model_upsample_smote_enn,
                                            newdata = data_test,
                                            type = "raw")

confusionMatrix(predictions_upsample_smote_enn_class,
                data_test$Class)

```

___

###### 3.2.4. Downsampling Methods -  Random Under-Sampling (RUS)

```{r 3.2 RUS, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


data_train_downsampled_ros <- downSample(
  x = subset(data_train, select = -Class),
  y = data_train$Class,
  yname = "Class")

table(data_train_downsampled_ros$Class)

model_downsample_ros <- train(Class ~ .,
                            data = data_train_downsampled_ros,
                            method = "glm",
                            trControl = train_control,
                            metric = "ROC")

predictions_downsample_ros <- predict(model_downsample_ros, newdata = data_test)
confusionMatrix(predictions_downsample_ros, data_test$Class)

```

___

##### 3.3. Cost-Sensitive Techniques

###### 3.3.1. Regular Random Forest

```{r 3.4 rf, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


#  Random Forest
rf <- train(Class ~ .,
                     data = data_train,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     num.trees = 100,
                     importance = 'permutation'
                    )

predictions_rf <- predict(rf,
                          newdata = data_test)
confusionMatrix(predictions_rf, data_test$Class)


```
___

###### 3.3.2. Weighted Random Forest

```{r 3.5 rf_weighted, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


# Calculate case weights
n_negative_train <- sum(data_train$Class == "negative")
n_positive_train <- sum(data_train$Class == "positive")

# positive is minority
model_weights <- c(
  negative = (n_negative_train + n_positive_train) / (2 * n_negative_train),
  positive = (n_negative_train + n_positive_train) / (2 * n_positive_train))


# Weighted Random Forest
weighted_rf <- train(Class ~ .,
                     data = data_train,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     class.weights = model_weights,
                     num.trees = 100,
                     importance = 'permutation'
                    )

predictions_weighted_rf <- predict(weighted_rf,
                                   newdata = data_test)
confusionMatrix(predictions_weighted_rf, data_test$Class)


```