---
title: "Comparing Methods for Handling Imbalanced Datasets"
author: "Hamed Helisaz"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(tidyverse)
library(imbalance)
library(GGally)
library(patchwork)
library(recipes)
library(themis)
library(smotefamily)
library(UBL)
library(unbalanced)
library(devtools)
library(caret)
library(formattable)
library(kableExtra)
library(pROC)
library(UBL)
set.seed(456)
```


### 1. Data Preparation 💾

This dataset is used for predicting the likelihood of a **cerebral stroke** based on various health and demographic factors. It is an **imbalanced classification problem**, making it suitable for evaluating models using metrics like **Precision**, **Recall**, **F1 Score**, **ROC-AUC**, and **PR-AUC**.

**Key Features Include:**
- Age, Hypertension, Heart Disease
- Average Glucose Level, BMI
- Smoking Status, Gender, Work Type, Residence Type
- Stroke (target variable)

**Class Imbalance:** The target variable (`stroke`) is highly imbalanced, with a small percentage of positive stroke cases.

**Source:** [Cerebral Stroke Prediction (Imbalanced Dataset) on Kaggle](https://www.kaggle.com/datasets/shashwatwork/cerebral-stroke-predictionimbalaced-dataset)

```{r 1.1 Load Data, echo = FALSE}

data_raw <- read.csv("data/data.csv")


factor_cols <- c("gender", "hypertension","heart_disease","ever_married",
                 "work_type","Residence_type","smoking_status")


data <- data_raw %>%
  mutate(Class = ifelse(stroke == 0, "negative","positive"))%>%
  select(-stroke)%>%
  mutate(Class = factor(Class, levels = c("negative","positive")))%>%
  select(-id)%>%
  mutate(across(all_of(factor_cols),as.factor))%>%
  ## Addressing NA
  mutate(smoking_status = ifelse(is.na(smoking_status), "Others",smoking_status))%>%
  filter(if_all(everything(), ~!is.na(.)))


best_threshold <- function(roc){
  
  roc_cords <- coords(roc,
                    ret = c("threshold", "ppv", "specificity", "sensitivity", "precision"),
                    transpose = FALSE)

  p <- roc_cords$precision
  s <- roc_cords$sensitivity
  roc_cords$f1 <- ifelse(p + s == 0, 0, 2 * p * s / (p + s))
  roc_cords <- roc_cords[!is.na(roc_cords$f1),]
  high_sens <- roc_cords [roc_cords['sensitivity']>= 0.9, ]
  best_thres <- high_sens[which.max(high_sens$f1), "threshold"]

  return(best_thres)
}


eval_fnc <- function(cm){

sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
precision   <- cm$byClass["Pos Pred Value"]
f1_score    <- 2 * (precision * sensitivity) / (precision + sensitivity)

print(cm$table)

cat("\nSensitivity:", round(sensitivity, 3))
cat("\nSpecificity:", round(specificity, 3))
cat("\nPrecision:   ", round(precision, 3))
cat("\nF1 Score:   ", round(f1_score, 3), "\n")

}

fontsize <- 8
basesize <- 8
```

___

### 2. Exploratory Data Analysis 🔍
#### 2.1. Initial Data Overview 📊

**Format**

A data frame with `r comma(nrow(data),0)` records and `r length(names(data))` columns; `r comma(nrow(data[data['Class'] == "positive",]),0)` comes with positive (minority) outcomes and `r comma(nrow(data[data['Class'] == "negative",]),0)` comes with negative (majority) outcomes (`r percent(nrow(data[data['Class'] == "positive",])/nrow(data),0)` imbalance). 

**Variables**

`r names(data)`

```{r 2.1 Initial Data Overview, echo = FALSE}

glimpse(data)

```


___

#### 2.2. Visualization
###### 2.2.1. Distribution of Classes

```{r 2.2 Class Distribution, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = 'center'}

ggplot(data, aes(x = Class, fill = Class)) +
  geom_bar() +
  labs(title = "Unbalanced Distribution", x = "Class", y = "Count") +
  theme_minimal(base_size = basesize) +
  scale_fill_manual(values = c("positive" = "skyblue", "negative" = "tomato"))+
  theme(axis.title.x = element_text(size = fontsize),
        axis.text.x = element_text(size = fontsize),
        axis.title.y = element_text(size = fontsize),
        axis.text.y = element_text(size = fontsize),
        plot.title = element_text(size = fontsize),
        legend.title = element_text(size = fontsize),
        legend.text = element_text(size = fontsize),
        legend.key.size =  unit(2, 'mm'),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7))

```

___

###### 2.2.2. Distribution of Predictors

```{r 2.2 Predictors Histograms, echo = FALSE, fig.width = 9, fig.height = 6, fig.align = 'center'}

numeric_predictors <- setdiff(names(data), c("Class",factor_cols))

plot_list <- list()

for (col_name in numeric_predictors) {
p <- ggplot(data, aes(x = .data[[col_name]])) +
    geom_histogram(aes(y = after_stat(density)), binwidth = NULL, fill = "lightblue", color = "black", alpha = 0.7) +
    geom_density(color = "darkblue", linewidth = 1) +
    labs(
      title = paste("Distribution of", col_name),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal()
   plot_list[[col_name]] <- p
}
   

combined_plot <- wrap_plots(plot_list, ncol = 2) 

print(combined_plot)

```

___

###### 2.2.3. Distribution of Predictors by Class - Histogram

```{r 2.2 Predictors Histograms by Class, echo = FALSE, fig.width = 9, fig.height = 6, fig.align = 'center'}


plot_list <- list()
class_colors <- c("positive" = "skyblue", "negative" = "tomato")

for (col_name in numeric_predictors) {
  p <- ggplot(data, aes(x = .data[[col_name]], fill = Class, color = Class)) +
    geom_density(alpha = 0.5, linewidth = 0.8) +
    scale_fill_manual(values = class_colors) +
    scale_color_manual(values = class_colors) +
    labs(
      title = paste("Distribution of", col_name, "by Class"),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal(base_size = fontsize -1) +
    theme(legend.position = "none")

  plot_list[[col_name]] <- p
}

combined_plot <- wrap_plots(plot_list, ncol = 2) +
                 plot_layout(guides = 'collect') &
                 theme(legend.position = 'right')

print(combined_plot)
```


###### 2.2.3. Distribution of Predictors by Class - Contingency Tables

```{r 2.2 Predictors Boxplot, echo = FALSE, results='asis'}

factor_predictors <- factor_cols  

tables_list <- list()

for (col_name in factor_predictors) {
  tbl_df <- data %>%
    group_by(!!sym(col_name), Class) %>%
    summarise(Count = n(), .groups = "drop") %>%
    group_by(!!sym(col_name)) %>%
    mutate(Percent = round(100 * Count / sum(Count), 1),
           Label = paste0(comma(Count), " (", Percent, "%)")) %>%
    select(!!sym(col_name), Class, Label) %>%
    pivot_wider(names_from = Class, values_from = Label) %>%
    mutate(across(everything(), ~replace_na(.x, "0 (0.0%)")))
  
  tables_list[[col_name]] <- kable(tbl_df, format = "html") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE, position = "left",
                  font_size = 12)
}


for (i in seq(1, length(tables_list), by = 4)) {
cat('<div style="display: flex; justify-content: flex-start; gap: 20px;">')

print(tables_list[[i]])

if ((i + 1) <= length(tables_list)) {
  print(tables_list[[i + 1]])
}

if ((i + 2) <= length(tables_list)) {
  print(tables_list[[i + 2]])
}

if ((i + 3) <= length(tables_list)) {
  print(tables_list[[i + 3]])
}

cat('</div><br>\n')
}
```

___

### 3. Methodology 🛠️

The following machine learning workflow is considered for all models:

**📦 Data Splitting:**  
The original dataset is divided into three parts:

- **Training set (≈56%)**: used to train the model.  
- **Validation set (≈24%)**: used to tune hyperparameters and select the optimal classification threshold.  
- **Test set (≈20%)**: held out and used **only once** for final model evaluation.  

This 3-way split ensures that model selection and threshold tuning do not bias the final reported performance.

**⚖️ Handling Class Imbalance:**  
All training datasets are augmented using **oversampling techniques** such as **SMOTE**, **ADASYN**, or **SMOTE-ENN** to address the severe class imbalance (positive class ≈1%).  
However, both the **validation** and **test** sets are **not augmented** and retain their original class distributions to reflect real-world scenarios.

**🎯 Model Training & Threshold Selection:**  
All models are trained to **maximize the F1 score**, while keeping sensitivity to **90 \%** which balances precision and recall, making it particularly suitable for highly imbalanced classification tasks. The classification **threshold** is tuned on the **validation set**. This tuned threshold is then applied to the test set for **final evaluation**, which is performed only once.

**📊 Evaluation Metrics:**  
Model performance is assessed using the following metrics:

- **F1 Score**: harmonic mean of precision and recall  
- **Precision (PPV)** and **Recall (Sensitivity)**  
- **Specificity** and **ROC-AUC** (where appropriate)  
- **PR-AUC** is also considered in some cases to better capture precision-recall trade-offs under imbalance  

This methodology ensures fair comparison and realistic performance estimation.

##### 3.1. Baseline Model (No Imbalance Treatment)

```{r 3.1 Baseline, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

train_control <- trainControl(method = "cv",
                              number = 5,
                              classProbs = TRUE,
                              summaryFunction = twoClassSummary,
                              savePredictions = "final")


test_index <- createDataPartition(data$Class, p = 0.8, list = FALSE)
data_train_valid <- data[test_index, ]
data_test <- data[-test_index, ]

train_index <- createDataPartition(data_train_valid$Class, p = 0.7, list = FALSE)
data_train <- data_train_valid[train_index, ]
data_valid <- data_train_valid[-train_index, ]

model_baseline <- train(Class ~ .,
                        data = data_train,
                        method = "glm",
                        trControl = train_control,
                        metric = "ROC")
                        ##"Recall", "Sensitivity", "Specificity"

print(model_baseline)
predictions_baseline <- predict(model_baseline, newdata = data_test)

table(data_train$Class)
cm_baseline <- confusionMatrix(predictions_baseline, data_test$Class, positive = "positive")
eval_fnc(cm_baseline)

```

___

##### 3.2. Resampling Techniques
###### 3.2.1. Upsampling Methods - Random Over-Sampling (ROS)

```{r 3.1 ROS, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

data_train_upsampled_ros <- upSample(
  x = subset(data_train, select = -Class),
  y = data_train$Class,
  yname = "Class")

table(data_train_upsampled_ros$Class)

model_upsample_ros <- train(Class ~ .,
                            data = data_train_upsampled_ros,
                            method = "glm",
                            trControl = train_control,
                            metric = "ROC")

predictions_upsample_ros <- predict(model_upsample_ros, newdata = data_test)

cm_upsample_ros <- confusionMatrix(predictions_upsample_ros, 
                                   data_test$Class, positive = "positive")
eval_fnc(cm_upsample_ros)

```

___

###### 3.2.2. Upsampling Methods -  SMOTE (Synthetic Minority Over-sampling Technique)

```{r 3.1 SMOTE, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


prepped_recipe_smote <- recipe(Class ~ ., data = data_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_smote(Class)%>%prep()

data_train_upsampled_smote <- juice(prepped_recipe_smote)
data_test_upsampled_smote <- recipe(Class ~ ., data = data_test) %>%
  step_dummy(all_nominal_predictors()) %>%
  prep() %>%
  bake(new_data = NULL)

table(data_train_upsampled_smote$Class)


model_upsample_smote <- train(Class ~ .,
                              data = data_train_upsampled_smote,
                              method = "glm",
                              family = "binomial",
                              trControl = train_control,
                              metric = "ROC")

print(model_upsample_smote)

predictions_upsample_smote_prob <- predict(model_upsample_smote, newdata = data_test_upsampled_smote, type = "prob")

predictions_upsample_smote_class <- predict(model_upsample_smote, newdata = data_test_upsampled_smote, type = "raw")

cm_upsample_smote <- confusionMatrix(predictions_upsample_smote_class, 
                                   data_test$Class, positive = "positive")
eval_fnc(cm_upsample_smote)
```

___

###### 3.2.3. Upsampling Methods -  SMOTE-ENN

```{r 3.1 SMOTE-ENN, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

prepped_recipe_smote_enn <- recipe(Class ~ ., data = data_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_smote(Class, over_ratio = 1)%>%prep()


data_train_upsampled_smote_enn <- juice(prepped_recipe_smote_enn)
data_test_upsampled_smote_enn <- recipe(Class ~ ., data = data_test) %>%
  step_dummy(all_nominal_predictors()) %>%
  prep() %>%
  bake(new_data = NULL)

data_valid_upsampled_smote_enn <- recipe(Class ~ ., data = data_valid) %>%
  step_dummy(all_nominal_predictors()) %>%
  prep() %>%
  bake(new_data = NULL)

data_train_upsampled_smote_enn$Class <- ifelse(
  data_train_upsampled_smote_enn$Class == "negative", 0, 1)

data_train_upsampled_smote_enn <- ubENN(X = data_train_upsampled_smote_enn
                    [, setdiff(names(data_train_upsampled_smote_enn), "Class")],
                    Y =  data_train_upsampled_smote_enn$Class)

data_train_upsampled_smote_enn_final <-
  cbind(data_train_upsampled_smote_enn$X,
        data.frame(Class = factor(
          ifelse(
          data_train_upsampled_smote_enn$Y == 0, 'negative','positive'),
          level = c('negative','positive'))))

table(data_train_upsampled_smote_enn_final$Class)

glm_upsample_smote_enn <- train(Class ~ .,
                              data = data_train_upsampled_smote_enn_final,
                              method = "glm",
                              family = "binomial",
                              trControl = train_control,
                              metric = "ROC")


predictions_upsample_smote_enn <- predict(glm_upsample_smote_enn,
                                            newdata = data_test_upsampled_smote_enn,
                                            type = "raw")

cm_upsample_smote_enn <- confusionMatrix(predictions_upsample_smote_enn, 
                                   data_test$Class, positive = "positive")
eval_fnc(cm_upsample_smote_enn)

```

___

###### 3.2.4. Downsampling Methods -  Random Under-Sampling (RUS)

```{r 3.2 RUS, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

data_train_downsampled_ros <- downSample(
  x = subset(data_train, select = -Class),
  y = data_train$Class,
  yname = "Class")

table(data_train_downsampled_ros$Class)

model_downsample_ros <- train(Class ~ .,
                            data = data_train_downsampled_ros,
                            method = "glm",
                            trControl = train_control,
                            metric = "ROC")

predictions_downsample_ros <- predict(model_downsample_ros, newdata = data_test)

cm_downsample_ros <- confusionMatrix(predictions_downsample_ros, 
                                   data_test$Class, positive = "positive")
eval_fnc(cm_downsample_ros)

```

___

##### 3.3. Cost-Sensitive Techniques

###### 3.3.1. Regular Random Forest

```{r 3.4 rf, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


#  Random Forest
rf <- train(Class ~ .,
                     data = data_train,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     num.trees = 100,
                     importance = 'permutation'
                    )

predictions_rf <- predict(rf,
                          newdata = data_test)

table(data_train$Class)

cm_rf <- confusionMatrix(predictions_rf, 
                                   data_test$Class, positive = "positive")
eval_fnc(cm_rf)

```
___

###### 3.3.2. Weighted Random Forest

```{r 3.5 rf_weighted, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


n_negative_train <- sum(data_train$Class == "negative")
n_positive_train <- sum(data_train$Class == "positive")

model_weights <- c(
  negative = (n_negative_train + n_positive_train) / (2 * n_negative_train),
  positive = (n_negative_train + n_positive_train) / (2 * n_positive_train))


# Weighted Random Forest
weighted_rf <- train(Class ~ .,
                     data = data_train,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     class.weights = model_weights,
                     num.trees = 100, ## Default is 500
                     importance = 'permutation'
                    )

best_model <- weighted_rf$bestTune
cv_preds <- weighted_rf$pred
cv_preds_best <- subset(cv_preds,
                        mtry == best_model$mtry)

roc_obj_cv <- roc(response = cv_preds_best$obs,
                  predictor = cv_preds_best$positive, 
                  levels = c("negative", "positive"))

roc_cords <- coords(roc_obj_cv)
## Fixing the sensitivity at 95%
high_sens <- roc_cords[roc_cords$sensitivity >= 0.95, ]
## Choosing the best specificity
best_thresh <- high_sens[which.max(high_sens$specificity), "threshold"]


probs <- predict(weighted_rf, newdata = data_test, type = "prob")[, "positive"]

pred_classes <- ifelse(probs >= best_thresh, "positive", "negative")

pred_classes <- factor(pred_classes, levels = c("negative", "positive"))

cm_weighted_rf <- confusionMatrix(pred_classes, data_test$Class, positive = "positive")


```

ROC is used to find an optimal threshold for classification. The threshold is supposed to maximize the specificity while keeping the sensitivity at the minimum of 95\%. That leads to choosing the threshold of `r round(best_thresh,2)`

```{r 3.5 rf_weighted_report, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

eval_fnc(cm_weighted_rf)

```


###### 3.3.3. Random Forest with SMOTE-ENN

```{r 3.6 rf_SMOTE_ENN, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

rf_smote_enn <- train(Class ~ .,
                     data = data_train_upsampled_smote_enn_final,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     num.trees = 100, ## Default is 500
                     importance = 'permutation'
                    )

probs_valid <- predict(rf_smote_enn, newdata = data_valid_upsampled_smote_enn , type = "prob")[, "positive"]

roc_obj_valid <- roc(response = data_valid$Class,
                     predictor = probs_valid,
                     levels = c("negative", "positive"))


best_thresh <- best_threshold(roc_obj_valid)

probs <- predict(rf_smote_enn, newdata = data_test_upsampled_smote_enn, type = "prob")[, "positive"]

pred_classes <- ifelse(probs >= best_thresh, "positive", "negative")

pred_classes <- factor(pred_classes, levels = c("negative", "positive"))

cm_rf_smote_enn <- confusionMatrix(pred_classes, data_test_upsampled_smote_enn$Class, positive = "positive")


```

ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.6 rf_SMOTE_ENN_report, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

eval_fnc(cm_rf_smote_enn)

```


###### 3.3.4. Random Forest with ADASYN

```{r 3.7 rf_ADASYN, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

data_train$Class <- as.factor(data_train$Class)

X <- model.matrix(Class ~ . - 1, data = data_train) 
X <- as.data.frame(X)
X$Class <- data_train$Class
## The reason for using model.matrix is to ensure all factors are well transformed 
## into integer numbers

data_train_adasyn <- AdasynClassif(Class ~ ., X)


rf_adasyn <- train(Class ~ .,
                     data = data_train_adasyn,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     num.trees = 100, ## Default is 500
                     importance = 'permutation'
                    )

X_valid <- model.matrix(Class ~ . - 1, data = data_valid)
X_valid <- as.data.frame(X_valid)
X_valid$Class <- data_valid$Class

probs_valid <- predict(rf_adasyn, newdata = X_valid, type = "prob")[, "positive"]

roc_obj_valid <- roc(response = X_valid$Class,
                     predictor = probs_valid,
                     levels = c("negative", "positive"))


best_thresh <- best_threshold(roc_obj_valid)


X_test <- model.matrix(Class ~ . - 1, data = data_test)
X_test <- as.data.frame(X_test)
X_test$Class <- data_test$Class

probs <- predict(rf_adasyn, newdata = X_test, type = "prob")[, "positive"]

pred_classes <- ifelse(probs >= best_thresh, "positive", "negative")

pred_classes <- factor(pred_classes, levels = c("negative", "positive"))

cm_rf_adasyn <- confusionMatrix(pred_classes, X_test$Class, positive = "positive")

```

Again, ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.7 rf_ADASYN_report, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

eval_fnc(cm_rf_adasyn)

```

