---
title: "Comparing Methods for Handling Imbalanced Datasets"
author: "Hamed Helisaz"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(tidyverse)
library(caret)
library(imbalance)
library(GGally)
library(patchwork)
library(recipes)
library(themis)
set.seed(456)
```


### 1. Data Preparation 💾

Loading `newthyroid1` data set from `imbalanced` library

```{r 1.1 Load Data, echo = FALSE}

data(newthyroid1)
thyroid_df_raw <- newthyroid1

thyroid_df <- thyroid_df_raw %>%
  mutate(Class = factor(Class, levels = c("negative","positive")))

fontsize <- 8
basesize <- 8
```

___

### 2. Exploratory Data Analysis 🔍
#### 2.1. Initial Data Overview 📊

**Format**

A data frame with 215 instances, 35 of which belong to the positive class, and 6 variables:

**Variables**

* **T3resin**: T3-resin uptake test, percentage. Discrete attribute.
* **Thyroxin**: Total Serum thyroxin as measured by the isotopic displacement method. Continuous attribute.
* **Triiodothyronine**: Total serum triiodothyronine as measured by radioimmuno assay. Continuous attribute.
* **Thyroidstimulating**: Basal thyroid-stimulating hormone (TSH) as measured by radioimmuno assay. Continuous attribute.
* **TSH\_value**: Maximal absolute difference of TSH value after injection of 200 micrograms of thyrotropin-releasing hormone as compared to the basal value. Continuous attribute.
* **Class**: Two possible classes: positive as hyperthyroidism, negative as non-hyperthyroidism.

```{r 2.1 Initial Data Overview, echo = FALSE}

glimpse(thyroid_df)

```


___

#### 2.2. Visualization
###### 2.2.1. Distribution of Classes

```{r 2.2 Class Distribution, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = 'center'}

ggplot(thyroid_df, aes(x = Class, fill = Class)) +
  geom_bar() +
  labs(title = "Unbalanced Distribution of Hyperthyroidism Status", x = "Hyperthyroidism Status", y = "Count") +
  theme_minimal(base_size = basesize) +
  scale_fill_manual(values = c("positive" = "skyblue", "negative" = "tomato"))+
  theme(axis.title.x = element_text(size = fontsize),
        axis.text.x = element_text(size = fontsize),
        axis.title.y = element_text(size = fontsize),
        axis.text.y = element_text(size = fontsize),
        plot.title = element_text(size = fontsize),
        legend.title = element_text(size = fontsize),
        legend.text = element_text(size = fontsize),
        legend.key.size =  unit(2, 'mm'),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7))

```

___

###### 2.2.2. Distribution of Predictors

```{r 2.2 Predictors Histograms, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

numeric_predictors <- setdiff(names(thyroid_df), "Class")

plot_list <- list()

for (col_name in numeric_predictors) {
p <- ggplot(thyroid_df, aes(x = .data[[col_name]])) +
    geom_histogram(aes(y = after_stat(density)), binwidth = NULL, fill = "lightblue", color = "black", alpha = 0.7) +
    geom_density(color = "darkblue", linewidth = 1) +
    labs(
      title = paste("Distribution of", col_name),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal()
   plot_list[[col_name]] <- p
}
   

combined_plot <- wrap_plots(plot_list, ncol = 2) 

print(combined_plot)

```

___

###### 2.2.3. Distribution of Predictors by Class

```{r 2.2 Predictors Histograms by Class, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

numeric_predictors <- setdiff(names(thyroid_df), "Class")

plot_list <- list()
class_colors <- c("positive" = "skyblue", "negative" = "tomato")

for (col_name in numeric_predictors) {
  p <- ggplot(thyroid_df, aes(x = .data[[col_name]], fill = Class, color = Class)) +
    geom_density(alpha = 0.5, linewidth = 0.8) + 
    scale_fill_manual(values = class_colors) +
    scale_color_manual(values = class_colors) +
    labs(
      title = paste("Distribution of", col_name, "by Class"),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal(base_size = fontsize -1) + 
    theme(legend.position = "none") 

  plot_list[[col_name]] <- p
}

combined_plot <- wrap_plots(plot_list, ncol = 2) + 
                 plot_layout(guides = 'collect') & 
                 theme(legend.position = 'right')

print(combined_plot)
```

___

### 3. Methodology 🛠️

##### 3.1. Baseline Model (No Imbalance Treatment)

```{r 3.1 Baseline, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

train_control <- trainControl(method = "cv",
                              number = 5, 
                              classProbs = TRUE,
                              summaryFunction = twoClassSummary, 
                              savePredictions = "final")


train_index <- createDataPartition(thyroid_df$Class, p = 0.7, list = FALSE)
thyroid_df_train <- thyroid_df[train_index, ]
thyroid_df_test <- thyroid_df[-train_index, ]

model_baseline <- train(Class ~ .,
                        data = thyroid_df_train,
                        method = "glm", 
                        trControl = train_control,
                        metric = "ROC") 
                        ##  "F1", "Recall", "Sensitivity", "Specificity"

print(model_baseline)
predictions_baseline <- predict(model_baseline, newdata = thyroid_df_test)

table(thyroid_df_train$Class)
confusionMatrix(predictions_baseline, thyroid_df_test$Class)

```

___

##### 3.2. Resampling Techniques
###### 3.2.1. Upsampling Methods - Random Over-Sampling (ROS)

```{r 3.1 ROS, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

thyroid_df_train_upsampled_ros <- upSample(x = subset(thyroid_df_train, 
                                                select = -Class),
                                     y = thyroid_df_train$Class,
                                     yname = "Class") 

table(thyroid_df_train_upsampled_ros$Class)

model_upsample_ros <- train(Class ~ .,
                            data = thyroid_df_train_upsampled_ros,
                            method = "glm", 
                            trControl = train_control,
                            metric = "ROC")

predictions_upsample_ros <- predict(model_upsample_ros, newdata = thyroid_df_test)
confusionMatrix(predictions_upsample_ros, thyroid_df_test$Class)

```

___

###### 3.2.2. Upsampling Methods -  SMOTE (Synthetic Minority Over-sampling Technique)

```{r 3.1 SMOTE, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}

smote_recipe <- recipe(Class ~ ., data = thyroid_df_train) %>%
  step_smote(Class, over_ratio = 1) 


thyroid_df_train_upsampled_smote <- juice(prep(smote_recipe))

table(thyroid_df_train_upsampled_smote$Class)


model_upsample_smote <- train(Class ~ .,
                              data = thyroid_df_train_upsampled_smote,
                              method = "glm",
                              family = "binomial",
                              trControl = train_control,
                              metric = "ROC")

print(model_upsample_smote)

predictions_upsample_smote_prob <- predict(model_upsample_smote, newdata = thyroid_df_test, type = "prob")

predictions_upsample_smote_class <- predict(model_upsample_smote, newdata = thyroid_df_test, type = "raw")

confusionMatrix(predictions_upsample_smote_class, 
                thyroid_df_test$Class)

```