---
title: "Comparing Methods for Handling Imbalanced Datasets"
author: "Hamed Helisaz"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(tidyverse)
library(imbalance)
library(GGally)
library(patchwork)
library(recipes)
library(themis)
library(smotefamily)
library(devtools)
library(caret)
library(formattable)
library(kableExtra)
library(pROC)
library(xgboost)
library(tidymodels)
library(xgboost)
library(torch)
set.seed(456)
```


### 1. Data Preparation 💾

This dataset is used for predicting the likelihood of a **cerebral stroke** based on various health and demographic factors. It is an **imbalanced classification problem**, making it suitable for evaluating models using metrics like **Precision**, **Recall**, **F1 Score**, **ROC-AUC**, and **PR-AUC**.

**Key Features Include:**
- Age, Hypertension, Heart Disease
- Average Glucose Level, BMI
- Smoking Status, Gender, Work Type, Residence Type
- Stroke (target variable)

**Class Imbalance:** The target variable (`stroke`) is highly imbalanced, with a small percentage of positive stroke cases.

**Source:** [Cerebral Stroke Prediction (Imbalanced Dataset) on Kaggle](https://www.kaggle.com/datasets/shashwatwork/cerebral-stroke-predictionimbalaced-dataset)

```{r 1.1 Load Data, echo = FALSE}

data_raw <- read.csv("data/data.csv")


factor_cols <- c("gender", "hypertension","heart_disease","ever_married",
                 "work_type","Residence_type","smoking_status")


data <- data_raw %>%
  mutate(Class = ifelse(stroke == 0, "negative","positive"))%>%
  select(-stroke)%>%
  mutate(Class = factor(Class, levels = c("negative","positive")))%>%
  select(-id)%>%
  mutate(across(all_of(factor_cols),as.factor))%>%
  ## Addressing NA
  mutate(smoking_status = ifelse(is.na(smoking_status), "Others",smoking_status))%>%
  filter(if_all(everything(), ~!is.na(.)))

### Find the best threshold based on the valuation criteria
best_threshold <- function(roc){
  
  roc_cords <- coords(roc,
                    ret = c("threshold", "ppv", "specificity", "sensitivity", "precision"),
                    transpose = FALSE)

  p <- roc_cords$precision
  s <- roc_cords$sensitivity
  roc_cords$f1 <- ifelse(p + s == 0, 0, 2 * p * s / (p + s))
  roc_cords <- roc_cords[!is.na(roc_cords$f1),]
  high_sens <- roc_cords
  # high_sens <- roc_cords [roc_cords['sensitivity']>= 0.9, ]
  best_thres <- high_sens[which.max(high_sens$f1), "threshold"]
  
  return(best_thres)
}
### Find the model + testing it to find the valuation criteria
evaluate_model <- function(model, valid_data, test_data) {
  
  is_torch_model <- inherits(model, "nn_module")
  is_xgb_model <- inherits(model, "xgb.Booster")

  predict_probs <- function(model, newdata) {
    if (is_torch_model) {
  x_mat <- torch_tensor(as.matrix(model.matrix(Class ~ . - 1, data = newdata)), 
                        dtype = torch_float())
  probs_valid <-as.numeric(as_array( model(x_mat)))
} else if (is_xgb_model) {
  x_mat <- model.matrix(Class ~ . - 1, data = newdata)
  probs_valid <- predict(model, x_mat)
} else {
  probs_valid <- predict(model, newdata = newdata, type = "prob")[, "positive"]
  
}
    
    return(probs_valid)
    }
  
  probs_valid <- predict_probs(model, valid_data)
  # ROC on validation
  roc_obj_valid <- roc(response = valid_data$Class,
                       predictor = probs_valid,
                       levels = c("negative", "positive"))
  
  # Find best threshold
  best_thresh <- best_threshold(roc_obj_valid)
  
  # Predict on test set
  probs_test <- predict_probs(model, test_data)
  
  # Convert probabilities to classes
  pred_classes <- ifelse(probs_test >= best_thresh, "positive", "negative")
  pred_classes <- factor(pred_classes, levels = c("negative", "positive"))
  
  # Confusion matrix
  cm_rf_smote_enn <- confusionMatrix(pred_classes, test_data$Class, positive = "positive")
  
  # Return results
  return(list(best_thresh = best_thresh, cm = cm_rf_smote_enn))
}
### Report the important valuation criteria that we care 
eval_fnc <- function(cm){

sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
precision   <- cm$byClass["Pos Pred Value"]
f1_score    <- 2 * (precision * sensitivity) / (precision + sensitivity)

print(cm$table)

cat("\nSensitivity:", round(sensitivity, 3))
cat("\nSpecificity:", round(specificity, 3))
cat("\nPrecision:   ", round(precision, 3))
cat("\nF1 Score:   ", round(f1_score, 3), "\n")

}

fontsize <- 8
basesize <- 8
```

___

### 2. Exploratory Data Analysis 🔍
#### 2.1. Initial Data Overview 📊

**Format**

A data frame with `r comma(nrow(data),0)` records and `r length(names(data))` columns; `r comma(nrow(data[data['Class'] == "positive",]),0)` comes with positive (minority) outcomes and `r comma(nrow(data[data['Class'] == "negative",]),0)` comes with negative (majority) outcomes (`r percent(nrow(data[data['Class'] == "positive",])/nrow(data),0)` imbalance). 

**Variables**

`r names(data)`

```{r 2.1 Initial Data Overview, echo = FALSE}

glimpse(data)

```


___

#### 2.2. Visualization
###### 2.2.1. Distribution of Classes

```{r 2.2 Class Distribution, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = 'center'}

ggplot(data, aes(x = Class, fill = Class)) +
  geom_bar() +
  labs(title = "Unbalanced Distribution", x = "Class", y = "Count") +
  theme_minimal(base_size = basesize) +
  scale_fill_manual(values = c("positive" = "skyblue", "negative" = "tomato"))+
  theme(axis.title.x = element_text(size = fontsize),
        axis.text.x = element_text(size = fontsize),
        axis.title.y = element_text(size = fontsize),
        axis.text.y = element_text(size = fontsize),
        plot.title = element_text(size = fontsize),
        legend.title = element_text(size = fontsize),
        legend.text = element_text(size = fontsize),
        legend.key.size =  unit(2, 'mm'),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7))

```

___

###### 2.2.2. Distribution of Predictors

```{r 2.2 Predictors Histograms, echo = FALSE, fig.width = 9, fig.height = 6, fig.align = 'center'}

numeric_predictors <- setdiff(names(data), c("Class",factor_cols))

plot_list <- list()

for (col_name in numeric_predictors) {
p <- ggplot(data, aes(x = .data[[col_name]])) +
    geom_histogram(aes(y = after_stat(density)), binwidth = NULL, fill = "lightblue", color = "black", alpha = 0.7) +
    geom_density(color = "darkblue", linewidth = 1) +
    labs(
      title = paste("Distribution of", col_name),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal()
   plot_list[[col_name]] <- p
}
   

combined_plot <- wrap_plots(plot_list, ncol = 2) 

print(combined_plot)

```

___

###### 2.2.3. Distribution of Predictors by Class - Histogram

```{r 2.2 Predictors Histograms by Class, echo = FALSE, fig.width = 9, fig.height = 6, fig.align = 'center'}


plot_list <- list()
class_colors <- c("positive" = "skyblue", "negative" = "tomato")

for (col_name in numeric_predictors) {
  p <- ggplot(data, aes(x = .data[[col_name]], fill = Class, color = Class)) +
    geom_density(alpha = 0.5, linewidth = 0.8) +
    scale_fill_manual(values = class_colors) +
    scale_color_manual(values = class_colors) +
    labs(
      title = paste("Distribution of", col_name, "by Class"),
      x = col_name,
      y = "Density"
    ) +
    theme_minimal(base_size = fontsize -1) +
    theme(legend.position = "none")

  plot_list[[col_name]] <- p
}

combined_plot <- wrap_plots(plot_list, ncol = 2) +
                 plot_layout(guides = 'collect') &
                 theme(legend.position = 'right')

print(combined_plot)
```


###### 2.2.3. Distribution of Predictors by Class - Contingency Tables

```{r 2.2 Predictors Boxplot, echo = FALSE, results='asis'}

factor_predictors <- factor_cols  

tables_list <- list()

for (col_name in factor_predictors) {
  tbl_df <- data %>%
    group_by(!!sym(col_name), Class) %>%
    summarise(Count = n(), .groups = "drop") %>%
    group_by(!!sym(col_name)) %>%
    mutate(Percent = round(100 * Count / sum(Count), 1),
           Label = paste0(comma(Count), " (", Percent, "%)")) %>%
    select(!!sym(col_name), Class, Label) %>%
    pivot_wider(names_from = Class, values_from = Label) %>%
    mutate(across(everything(), ~replace_na(.x, "0 (0.0%)")))
  
  tables_list[[col_name]] <- kable(tbl_df, format = "html") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE, position = "left",
                  font_size = 12)
}


for (i in seq(1, length(tables_list), by = 4)) {
cat('<div style="display: flex; justify-content: flex-start; gap: 20px;">')

print(tables_list[[i]])

if ((i + 1) <= length(tables_list)) {
  print(tables_list[[i + 1]])
}

if ((i + 2) <= length(tables_list)) {
  print(tables_list[[i + 2]])
}

if ((i + 3) <= length(tables_list)) {
  print(tables_list[[i + 3]])
}

cat('</div><br>\n')
}
```

___

### 3. Methodology 🛠️

The following machine learning workflow is considered for all models:

**📦 Data Splitting:**  
The original dataset is divided into three parts: I used **stratified splits** to ensure the minority class is represented in all subsets. 

- **Training set (≈56%)**: used to train the model.  
- **Validation set (≈24%)**: used to tune hyperparameters and select the optimal classification threshold.  
- **Test set (≈20%)**: held out and used **only once** for final model evaluation.  

This 3-way split ensures that model selection and threshold tuning do not bias the final reported performance.

```{r  data_split, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

test_index <- createDataPartition(data$Class, p = 0.8, list = FALSE)
data_train_valid <- data[test_index, ]
data_test <- data[-test_index, ]

train_index <- createDataPartition(data_train_valid$Class, p = 0.7, list = FALSE)
data_train <- data_train_valid[train_index, ]
data_valid <- data_train_valid[-train_index, ]

```


**⚖️ Handling Class Imbalance:**  
All training datasets are augmented using **oversampling techniques** such as **SMOTE**, **ADASYN**, or **SMOTE-ENN** to address the severe class imbalance (positive class ≈1%).  
However, both the **validation** and **test** sets are **not augmented** and retain their original class distributions to reflect real-world scenarios.

**🎯 Model Training & Threshold Selection:**  
All models are trained to **maximize the F1 score**, which balances precision and recall, making it particularly suitable for highly imbalanced classification tasks. The classification **threshold** is tuned on the **validation set**. This tuned threshold is then applied to the test set for **final evaluation**, which is performed only once.

**📊 Evaluation Metrics:**  
Model performance is assessed using the following metrics:

- **F1 Score**: harmonic mean of precision and recall  
- **Precision (PPV)** and **Recall (Sensitivity)**  
- **Specificity** and **ROC-AUC** (where appropriate)  
- **PR-AUC** is also considered in some cases to better capture precision-recall trade-offs under imbalance  

This methodology ensures fair comparison and realistic performance estimation.

___

#### 3.1. Generalized Linear Models (GLM) 🔢

In this section, we evaluate **logistic regression (GLM)** for handling the imbalanced dataset.  
We explore the following strategies:

1. **Baseline logistic regression** without any class imbalance treatment.  
2. **Resampling approaches**: Random Over-Sampling (ROS), SMOTE, SMOTE-ENN, Random Under-Sampling (RUS)  

The classification **threshold is tuned on the validation set** to maximize the **F1 score**,  
ensuring that the model achieves a balance between **precision** and **recall** while accounting for the class imbalance.


##### 3.1.1 Baseline GLM

```{r  GLM_Baseline, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

train_control <- trainControl(method = "cv",
                              number = 5,
                              classProbs = TRUE,
                              summaryFunction = twoClassSummary,
                              savePredictions = "final")


model_baseline <- train(Class ~ .,
                        data = data_train,
                        method = "glm",
                        trControl = train_control,
                        metric = "ROC")
                        ##"Recall", "Sensitivity", "Specificity"

print(model_baseline)
predictions_baseline <- predict(model_baseline, newdata = data_test)

table(data_train$Class)
cm_baseline <- confusionMatrix(predictions_baseline, data_test$Class, positive = "positive")
eval_fnc(cm_baseline)

```



##### 3.1.2 GLM + ROS

```{r GLM_ROS, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

data_train_ros <- upSample(
  x = subset(data_train, select = -Class),
  y = data_train$Class,
  yname = "Class")


table(data_train_ros$Class)

glm_ros <- train(Class ~ .,
                            data = data_train_ros,
                            method = "glm",
                            trControl = train_control,
                            metric = "ROC")

glm_ros_eval <- evaluate_model(glm_ros, 
               data_valid,
               data_test)

cm_glm_ros <- glm_ros_eval$cm

best_thresh <- glm_ros_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_glm_ros)

```
___

##### 3.1.3 GLM + SMOTE

```{r GLM_SMOTE, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

prepped_recipe_smote <- recipe(Class ~ ., data = data_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_smote(Class)%>%prep()

data_train_smote <- juice(prepped_recipe_smote)

data_valid_smote <- recipe(Class ~ ., data = data_valid) %>%
  step_dummy(all_nominal_predictors()) %>%
  prep() %>%
  bake(new_data = NULL)

data_test_smote <- recipe(Class ~ ., data = data_test) %>%
  step_dummy(all_nominal_predictors()) %>%
  prep() %>%
  bake(new_data = NULL)



glm_smote <- train(Class ~ .,
                              data = data_train_smote,
                              method = "glm",
                              family = "binomial",
                              trControl = train_control,
                              metric = "ROC")

glm_smote_eval <- evaluate_model(glm_smote, 
               data_valid_smote,
               data_test_smote)

cm_glm_smote <- glm_smote_eval$cm

best_thresh <- cm_glm_smote$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_glm_smote)

```
___

##### 3.1.4 GLM + SMOTE-ENN

```{r GLM_SMOTE-ENN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

prepped_recipe_smote_enn <- recipe(Class ~ ., data = data_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_smote(Class, over_ratio = 1)%>%prep()


data_train_smote_enn_start <- juice(prepped_recipe_smote_enn)
data_test_smote_enn <- recipe(Class ~ ., data = data_test) %>%
  step_dummy(all_nominal_predictors()) %>%
  prep() %>%
  bake(new_data = NULL)

data_valid_smote_enn <- recipe(Class ~ ., data = data_valid) %>%
  step_dummy(all_nominal_predictors()) %>%
  prep() %>%
  bake(new_data = NULL)

data_train_smote_enn_start$Class <- ifelse(
  data_train_smote_enn_start$Class == "negative", 0, 1)

data_train_smote_enn_start <- ubENN(X = data_train_smote_enn_start
                    [, setdiff(names(data_train_smote_enn_start), "Class")],
                    Y =  data_train_smote_enn_start$Class)

data_train_smote_enn <-
  cbind(data_train_smote_enn_start$X,
        data.frame(Class = factor(
          ifelse(
          data_train_smote_enn_start$Y == 0, 'negative','positive'),
          level = c('negative','positive'))))

table(data_train_smote_enn$Class)

glm_smote_enn <- train(Class ~ .,
                              data = data_train_smote_enn,
                              method = "glm",
                              family = "binomial",
                              trControl = train_control,
                              metric = "ROC")

glm_smote_enn_eval <- evaluate_model(glm_smote_enn, 
               data_valid_smote_enn,
               data_test_smote_enn)

cm_glm_smote_enn <- glm_smote_enn_eval$cm

best_thresh <- glm_smote_enn_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_glm_smote_enn)

```


___

##### 3.1.5 GLM + RUS 

```{r GLM_RUS, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

data_train_rus <- downSample(
  x = subset(data_train, select = -Class),
  y = data_train$Class,
  yname = "Class")

table(data_train_rus$Class)

glm_rus <- train(Class ~ .,
                            data = data_train_rus,
                            method = "glm",
                            trControl = train_control,
                            metric = "ROC")

glm_rus_eval <- evaluate_model(glm_rus, 
               data_valid,
               data_test)

cm_rus <- glm_rus_eval$cm

best_thresh <- glm_rus_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_rus)

```

___

#### 3.2 Random Forest 🌲

Random Forest is a robust ensemble method that often performs well with unbalanced data due to its ability to handle non-linear relationships.  
In this section, we test different **Random Forest variations**:

1. **Baseline logistic regression** without any class imbalance treatment.  
2. **Resampling approaches**: SMOTE-ENN and ADASYN
3. **Cost-Sensitive Weights**

The classification **threshold is tuned on the validation set** to maximize the **F1 score**,  
ensuring that the model achieves a balance between **precision** and **recall** while accounting for the class imbalance.

##### 3.2.1. Baseline RF

```{r 3.4 rf, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}


#  Random Forest
rf <- train(Class ~ .,
                     data = data_train,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     num.trees = 100,
                     importance = 'permutation'
                    )

rf_eval <- evaluate_model(rf, 
               data_valid,
               data_test)

cm_rf <- rf_eval$cm

best_thresh <- rf_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.6 rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_rf)

```

___


##### 3.2.2 RF + SMOTE-ENN

```{r rf_SMOTE_ENN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

rf_smote_enn <- train(Class ~ .,
                     data = data_train_smote_enn_final,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     num.trees = 100, ## Default is 500
                     importance = 'permutation'
                    )

rf_smote_enn_eval <- evaluate_model(rf_smote_enn, 
               data_valid_smote_enn,
               data_test_smote_enn)

cm_rf_smote_enn <- glm_upsample_smote_enn_eval$cm

best_thresh <- glm_upsample_smote_enn_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_rf_smote_enn)

```

___


##### 3.2.3 RF + ADASYN

```{r rf_ADASYN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

data_train$Class <- as.factor(data_train$Class)

X_train <- model.matrix(Class ~ . - 1, data = data_train) 
X_train <- as.data.frame(X_train)
X_train$Class <- data_train$Class
## The reason for using model.matrix is to ensure all factors are well transformed 
## into integer numbers

data_train_adasyn <- AdasynClassif(Class ~ ., X_train)


rf_adasyn <- train(Class ~ .,
                     data = data_train_adasyn,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     num.trees = 100, ## Default is 500
                     importance = 'permutation'
                    )

X_valid <- model.matrix(Class ~ . - 1, data = data_valid)
X_valid <- as.data.frame(X_valid)
X_valid$Class <- data_valid$Class


X_test <- model.matrix(Class ~ . - 1, data = data_test)
X_test <- as.data.frame(X_test)
X_test$Class <- data_test$Class

rf_adasyn_eval <- evaluate_model(rf_adasyn, 
               X_valid,
               X_test)

cm_rf_adasyn <- rf_adasyn_eval$cm

best_thresh <- rf_adasyn_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.6 rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_rf_adasyn)

```

##### 3.2.4 Weighted RF

```{r rf_weighted, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center'}


n_negative_train <- sum(data_train$Class == "negative")
n_positive_train <- sum(data_train$Class == "positive")

model_weights <- c(
  negative = (n_negative_train + n_positive_train) / (2 * n_negative_train),
  positive = (n_negative_train + n_positive_train) / (2 * n_positive_train))


# Weighted Random Forest
weighted_rf <- train(Class ~ .,
                     data = data_train,
                     method = "ranger",
                     trControl = train_control,
                     metric = "ROC",
                     class.weights = model_weights,
                     num.trees = 100, ## Default is 500
                     importance = 'permutation'
                    )

weighted_rf_eval <- evaluate_model(weighted_rf, 
               data_valid,
               data_test)

cm_weighted_rf <- weighted_rf_eval$cm

best_thresh <- weighted_rf_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.6 rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_weighted_rf)

```

___

#### 3.3 Boosting (XGBoost) ⚡

In this section, we explore the **XGBoost** algorithm, a powerful gradient boosting technique, for handling the severe class imbalance.

We compare the following strategies:
1. **Baseline logistic regression** without any class imbalance treatment. 
2. **Resampling approaches**: **SMOTE**, **SMOTE-ENN**, and **ADASYN**.  
3. **Cost-Sensitive Weights**

As with previous models, the **classification threshold** is tuned on the **validation set** to maximize the **F1 score**, ensuring a balance between precision and recall.

##### 3.3.1 Baseline XGBoost 

```{r XGBoost, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train <- model.matrix(Class ~ . - 1, data = data_train)
y_train <- ifelse(data_train$Class == "positive", 1, 0)

xgb <- xgboost(
  data = x_train,
  label = y_train,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc"
)

xgb_eval <- evaluate_model(xgb, 
               data_valid,
               data_test)

cm_xgb <- xgb_eval$cm

best_thresh <- xgb_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r XGBoost_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_xgb)

```

___



##### 3.3.2 XGBoost + SMOTE 

```{r XGBOOST_SMOTE, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train_smote <- model.matrix(Class ~ . - 1, data = data_train_smote)
y_train_smote <- ifelse(data_train_smote$Class == "positive", 1, 0)

xgb_smote <- xgboost(
  data = x_train_smote,
  label = y_train_smote,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc"
)

xgb_smote_eval <- evaluate_model(xgb_smote, 
               data_valid_smote,
               data_test_smote)

cm_xgb_smote <- xgb_smote_eval$cm

best_thresh <- xgb_smote_eval$best_thresh 


```

Again, ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.7 XGBOOST_SMOTE_report , echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_xgb_smote)

```


##### 3.3.3 XGBoost + SMOTE-ENN

```{r XGBOOST_SMOTE_ENN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train_smote_enn <- model.matrix(Class ~ . - 1, data = data_train_smote_enn)
y_train_smote_enn <- ifelse(data_train_smote_enn$Class == "positive", 1, 0)

xgb_smote_enn <- xgboost(
  data = x_train_smote_enn,
  label = y_train_smote_enn,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc"
)

xgb_smote_enn_eval <- evaluate_model(xgb_smote_enn, 
               data_valid_smote_enn,
               data_test_smote_enn)

cm_xgb_smote_enn <- xgb_smote_enn_eval$cm

best_thresh <- xgb_smote_enn_eval$best_thresh 


```

Again, ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.7 XGBOOST_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_xgb_smote_enn)

```

##### 3.3.4 XGBoost + ADASYN


```{r XGBoost_ADASYN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train_adasyn <- model.matrix(Class ~ . - 1, data = data_train_adasyn)
y_train_adasyn <- ifelse(data_train_adasyn$Class == "positive", 1, 0)


xgb_adasyn <- xgboost(
  data = x_train_adasyn,
  label = y_train_adasyn,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc"
)

xgb_adasyn_eval <- evaluate_model(xgb_adasyn, 
               X_valid,
               X_test)

cm_xgb_adasyn <- xgb_adasyn_eval$cm

best_thresh <- xgb_adasyn_eval$best_thresh 


```

Again, ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.7 XGBOOST_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_xgb_adasyn)

```

##### 3.3.5 Weighted XGBoost

```{r XGBoost_weighted, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}


weight_negative <- (n_negative_train + n_positive_train) / (2 * n_negative_train)
weight_positive <- (n_negative_train + n_positive_train) / (2 * n_positive_train)

# Create a vector of weights per row
row_weights <- ifelse(y_train == 1, model_weights[['positive']], model_weights[['negative']])


weighted_xgb <- xgboost(
  data = x_train,
  label = y_train,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc",
  weight = row_weights
)

weighted_xgb_eval <- evaluate_model(weighted_xgb, 
               data_valid,
               data_test)

cm_weighted_xgb <- weighted_xgb_eval$cm

best_thresh <- weighted_xgb_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.6 rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_weighted_xgb)

```


#### 3.4  Neural Networks (NN) 🧠

In this section, we explore **feed-forward Neural Networks (NN)** for the same task. Neural networks can capture complex patterns but are also sensitive to class imbalance.

We compare the following strategies:
1. **Baseline logistic regression** without any class imbalance treatment. 
2. **Resampling approaches**: **SMOTE**, **SMOTE-ENN**, and **ADASYN**.  
3. **Cost-Sensitive Weights**

As with previous models, the **classification threshold** is tuned on the **validation set** to maximize the **F1 score**, ensuring a balance between precision and recall.

##### 3.4.1 Baseline NN 

```{r NN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train <- torch_tensor(as.matrix(model.matrix(Class ~ . - 1, data = data_train)), 
                        dtype = torch_float())
y_train <- torch_tensor(ifelse(data_train$Class == "positive", 1, 0), 
                        dtype = torch_float())$unsqueeze(2)

nn_define <- nn_module(
  "Net",
  initialize = function(input_dim) {
    self$fc1 <- nn_linear(input_dim, 16)
    self$dropout <- nn_dropout(p = 0.3)
    self$fc2 <- nn_linear(16, 8)
    self$fc3 <- nn_linear(8, 1)
  },
  forward = function(x) {
    x %>%
      self$fc1() %>%
      nnf_relu() %>%
      self$dropout() %>%
      self$fc2() %>%
      nnf_relu() %>%
      self$fc3() %>%
      nnf_sigmoid()
  }
)

nn <- nn_define(input_dim = ncol(x_train))

optimizer <- optim_adam(nn$parameters, lr = 0.001)
loss_fn <- function(input, target) {
  nnf_binary_cross_entropy(input, target, weight = ifelse(target == 1, 5, 1))  # example weights
}

for (epoch in 1:100) {
  nn$train()
  optimizer$zero_grad()
  output <- nn(x_train)
  loss <- loss_fn(output, y_train)
  loss$backward()
  optimizer$step()

  if (epoch %% 10 == 0) {
    cat(sprintf("Epoch %d, Loss: %.4f\n", epoch, loss$item()))
  }
}

nn_eval <- evaluate_model(nn, 
               data_valid,
               data_test)

cm_nn <- nn_eval$cm

best_thresh <- nn_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r XGBoost_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_nn)

```

___



##### 3.3.2 XGBoost + SMOTE 

```{r XGBOOST_SMOTE, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train_smote <- model.matrix(Class ~ . - 1, data = data_train_smote)
y_train_smote <- ifelse(data_train_smote$Class == "positive", 1, 0)

xgb_smote <- xgboost(
  data = x_train_smote,
  label = y_train_smote,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc"
)

xgb_smote_eval <- evaluate_model(xgb_smote, 
               data_valid_smote,
               data_test_smote)

cm_xgb_smote <- xgb_smote_eval$cm

best_thresh <- xgb_smote_eval$best_thresh 


```

Again, ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.7 XGBOOST_SMOTE_report , echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_xgb_smote)

```


##### 3.3.3 XGBoost + SMOTE-ENN

```{r XGBOOST_SMOTE_ENN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train_smote_enn <- model.matrix(Class ~ . - 1, data = data_train_smote_enn)
y_train_smote_enn <- ifelse(data_train_smote_enn$Class == "positive", 1, 0)

xgb_smote_enn <- xgboost(
  data = x_train_smote_enn,
  label = y_train_smote_enn,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc"
)

xgb_smote_enn_eval <- evaluate_model(xgb_smote_enn, 
               data_valid_smote_enn,
               data_test_smote_enn)

cm_xgb_smote_enn <- xgb_smote_enn_eval$cm

best_thresh <- xgb_smote_enn_eval$best_thresh 


```

Again, ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.7 XGBOOST_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_xgb_smote_enn)

```

##### 3.3.4 XGBoost + ADASYN


```{r XGBoost_ADASYN, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

x_train_adasyn <- model.matrix(Class ~ . - 1, data = data_train_adasyn)
y_train_adasyn <- ifelse(data_train_adasyn$Class == "positive", 1, 0)


xgb_adasyn <- xgboost(
  data = x_train_adasyn,
  label = y_train_adasyn,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc"
)

xgb_adasyn_eval <- evaluate_model(xgb_adasyn, 
               X_valid,
               X_test)

cm_xgb_adasyn <- xgb_adasyn_eval$cm

best_thresh <- xgb_adasyn_eval$best_thresh 


```

Again, ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.7 XGBOOST_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_xgb_adasyn)

```

##### 3.3.5 Weighted XGBoost

```{r XGBoost_weighted, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}


weight_negative <- (n_negative_train + n_positive_train) / (2 * n_negative_train)
weight_positive <- (n_negative_train + n_positive_train) / (2 * n_positive_train)

# Create a vector of weights per row
row_weights <- ifelse(y_train == 1, model_weights[['positive']], model_weights[['negative']])


weighted_xgb <- xgboost(
  data = x_train,
  label = y_train,
  objective = "binary:logistic",
  nrounds = 100,
  eval_metric = "auc",
  weight = row_weights
)

weighted_xgb_eval <- evaluate_model(weighted_xgb, 
               data_valid,
               data_test)

cm_weighted_xgb <- weighted_xgb_eval$cm

best_thresh <- weighted_xgb_eval$best_thresh 

```


ROC is used to find an optimal threshold for classification. This time, the threshold is `r round(best_thresh,2)`

```{r 3.6 rf_SMOTE_ENN_report, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

eval_fnc(cm_weighted_xgb)

```

